\documentclass{article}

% Language setting Replace `english' with e.g. `spanish' to change the document
% language
\usepackage[english]{babel}

% Set page size and margins Replace `letterpaper' with `a4paper' for UK/EU
% standard size
\usepackage[letterpaper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

% Useful packages
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage[pdftex]{graphicx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{todonotes}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{xcolor}
\usepackage{tabularray}
\usepackage{xspace} % For \xspace command
\usepackage{tikz}
\usetikzlibrary{shadings,positioning,overlay-beamer-styles}
\usepackage{xifthen}
\usepackage{ulem}
\usepackage{amsthm}
\usepackage{cleveref}

\newcommand{\YC}[1]{\textcolor{red}{YC: #1}}
\newcommand{\TG}[2][]{\ifthenelse{\equal{#1}{done}}{[\textcolor{teal}{$\blacksquare$}]~\textcolor{blue!60}{\textsc{From
Tristan:} #2}}{\textcolor{blue}{\textsc{From Tristan:} #2}}}


\providecommand{\includeappendix}{true}
\newcommand{\0}{\mspace{9mu}}
\newcommand{\navr}[0]{$\nu_{\mathrm{npv}}$\xspace}
\newcommand{\Var}[0]{\mathrm{Var}}

\title{The practical impact of numerical variability on structural MRI measures of Parkinson's disease}

\author{Yohan Chatelain, Andrzej Sokołowski, Madeleine Sharp, Jean-Baptiste Poline, Tristan Glatard}

\begin{document}

\maketitle

\begin{abstract}
    Numerical variability is rarely quantified in neuroimaging despite many
    biomarkers relying on subtle morphometric differences across individuals. We
    instrumented FreeSurfer, a widely used neuroimaging pipeline, to simulate
    numerical differences across computational environments, and used it to
    measure numerical variability in MRI analyses of Parkinson's disease
    patients and controls. In multiple cortical and subcortical regions,
    numerical variation reached nearly one-third of the population variability,
    altering statistical conclusions about group differences and clinical
    associations. To address this, we developed a practical tool that estimates
    the Numerical-Population Variability Ratio (NPVR), enabling researchers to
    assess the impact of numerical noise in existing studies. Applying
    NPVR-based thresholding to ENIGMA brain maps showed that many small regional
    effects fall below typical numerical noise thresholds \TG{Note to self:
        re-evaluate this (strong) claim after reading the results} for standard
    study sizes \TG{not sure why study sizes are mentioned here since ENIGMA
        maps were obtained for specific sample sizes}. Our tool provides a practical
    approach to quantify the effect of numerical variability, improving the
    robustness and reproducibility of neuroimaging analyses.
\end{abstract}

\input{main}

\include{methods}

\section{Data Availability}
The data that support the findings of this study are available from the
Parkinson's Progression Markers Initiative (PPMI) database
(www.ppmi-info.org/access-data-specimens/download-data), but restrictions apply
to the availability of these data, which were used under license for the
current study, and so are not publicly available. Data are however available
from the authors upon reasonable request and with permission of the PPMI.

\section{Code Availability}

All MCA instrumentation scripts, FreeSurfer build instructions and analysis
notebooks are available at [GitHub URL to be inserted]. Exact commit hashes are
archived on Zenodo (DOI [to be added]) to ensure bit-level reproducibility.

\section{Acknowledgements}

The analyses were conducted on the Virtual Imaging
Platform~\cite{glatard2012virtual}, which utilizes resources provided by the
Biomed virtual organization within the European Grid Infrastructure (EGI). We
extend our gratitude to Sorina Pop from CREATIS, Lyon, France, for her support.
\TG{acknowledge MJFF project LivingPark}

\bibliographystyle{plain}
\bibliography{paper}

\ifthenelse{\equal{\includeappendix}{true}}{
    \include{appendix}
}{}

\end{document}
